\large Testing cosmic isotropy and homogeneity \\ with the all-sky \textsl{Quaia} sample of quasars

### 2023-07-01 paper/main.tex
title, abstract, intro

% text definitions
\shortauthors{hogg and storey-fisher}
\shorttitle{testing isotropy and homogeneity}
\newcommand{\documentname}{\textsl{Article}}
\newcommand{\sectionname}{Section}
\newcommand{\secref}[1]{\sectionname~\ref{#1}}
\newcommand{\figref}[1]{\figurename~\ref{#1}}
\newcommand{\abby}[1]{\textbf{Abby says: #1}}

@@

\title{%
Testing cosmic isotropy and homogeneity
with \textsl{Gaia} quasars}
\author[0000-0003-2866-9403]{David W. Hogg}
\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University}
\affil{Max-Planck-Institut f\"ur Astronomie}
\affil{Flatiron Institute, a division of the Simons Foundation}

@@

\date{2022 August}

\begin{abstract}\noindent
All cosmological models in general relativity are built on the assumption of large-scale isotropy and homogeneity.
These assumptions are testable.
Tests of cosmic isotropy and homogeneity are also strong tests of the quality of data or catalogs and their calibration.
Test precision increases as observations cover more solid angle, become more sensitive to distant objects, and are made with more uniform data.
The ESA \textsl{Gaia} Mission is primarily concerned with stars in the Milky Way, but it also observes millions of quasars, all sky (modulo Galactic extinction), to redshifts of around $4$.
It has produced the largest-comoving-volume public quasar catalog to date.
Here we use a carefully curated subsample of ZZZ \textsl{Gaia} quasars to show that the angular distribution, redshift distribution as a function of sky position, amplitude of clustering as a function of sky position, and counts of pairs as a function of separation (fractal dimension), are all consistent with large-scale isotropy and homogeneity at the precision possible given the size of the catalog and the amplitude of the large-scale structure.
Roughly speaking, the results can be summarized by saying that they show the Universe to be isotropic to better than XXX percent and homogeneous to better than YYY percent.
There is no sign of any North--South power asymmetry.
\abby{Is this the same hemispherical power asymmetry claimed from the CMB?}
The isotropy results can also be interpreted as a strong test of the spectrophotometric uniformity of the \textsl{Gaia} Catalog.
\end{abstract}

@@

Another reason to make new tests of isotropy and homogeneity is that data samples keep getting larger in volume and number, permitting more and more sensitive tests.
Here we make use of the brand-new ESA \textsl{Gaia} DR3 data (CITE), which includes hundreds of thousands or even millions of quasars, all-sky (CITE).
This data set is not the largest cosmological data set in terms of number of tracers, but it is the largest-ever data set in terms of comoving volume, spanning some XXX $h^{-3}\,\Gpc^3$ (comoving), depending on how you want to define the sample volume.
Recent tests of homogeneity include ZZZ and ZZZ; the samples used in those studies were smaller by a factor of QQQ and WWW (approximately) in comoving volume, relative to the quasar catalog we use here.

A final but equally important reason to test isotropy and homogeneity is that these tests provide a very strong validation of the quality and uniformity of the data.
Indeed, the first work on homogeneity with the Sloan Digital Sky Survey data (\citealt{hogg05}) was performed to test the calibration and stability of the data in preparation for the measurement of the baryon acoustic feature (\citealt{baf}).
The work presented here represents a first strong test of the calibration of the \textsl{Gaia} quasar sample and associated photometry and spectroscopy; it has implications for the accuracy and uniformity of the \textsl{Gaia} data more generally.

\section{Data}\label{sec:data}\noindent
Hogg say: Make sure we cite the CU5 papers and XP papers here and anything that helped make the extragalactic and quasar samples.

@@

HOGG: DOES \cite{ksf} explain how the random catalog points get redshifts? Or are we doing that here?

@@

HOGG: \figref{fig:cumulative} strongly suggests that the fractal dimension $D$ is 3.
Our best-fit value is ... WHAT?

@@

\section{Discussion}\label{sec:discuss}\noindent
HOGG: We found what we expected to find, but at very good precision.

HOGG: How to summarize these results?

HOGG: ESA \textsl{Gaia} rocks!

\begin{acknowledgements}
It is a pleasure to thank
  Anna-Christina Eilers (MIT),
  Hans-Walter Rix (MPIA), and
  Abby Williams (NYU, Caltech) for valuable discussions. KSF: OTHERS?
This project was developed in part at the 2022 Gaia F\^ete, hosted by the Center for Computational Astrophysics at the Flatiron Institute.
This project was also developed in part at the Gaia Hike, a workshop hosted by the University of British Columbia and the Canadian Institute for Theoretical Astrophysics in 2022 June.

HOGG: Be sure to put in the full Gaia acknowlegement.

HOGG: Did we use 2MASS? If so, fix the text to capture it above and below.
\end{acknowledgements}

>>>>

% text definitions
\shortauthors{hogg and storey-fisher}
\shorttitle{testing isotropy and homogeneity}
\newcommand{\documentname}{\textsl{Article}}
\newcommand{\sectionname}{Section}
\newcommand{\secref}[1]{\sectionname~\ref{#1}}
\newcommand{\figref}[1]{\figurename~\ref{#1}}
\newcommand{\abby}[1]{\textbf{Abby says: #1}}
\newcommand{\hogg}[1]{\textbf{Hogg says: #1}}

@@

\title{\large%
Testing cosmic isotropy and homogeneity \\ with the all-sky \textsl{Quaia} sample of quasars}
\author[0000-0003-2866-9403]{David W. Hogg}
\affil{Center for Cosmology and Particle Physics, Department of Physics, New York University}
\affil{Max-Planck-Institut f\"ur Astronomie}
\affil{Flatiron Institute, a division of the Simons Foundation}

@@

\date{2023 July}

\begin{abstract}\noindent
All cosmological models in general relativity are built on the assumption of large-scale isotropy and homogeneity.
These assumptions are testable.
Tests of cosmic isotropy and homogeneity are also strong tests of the quality of data or catalogs and their calibration.
Test precision increases as observations cover more solid angle, become more sensitive to distant objects, and are made with more uniform data.
The ESA \textsl{Gaia} Mission is primarily concerned with stars in the Milky Way, but it also observes millions of quasars, all sky (modulo Galactic extinction), to redshifts of around $4$.
It has produced the largest-comoving-volume public quasar catalog to date.
Here we use the $755\,850$ quasars in the \textsl{Quaia} $G<20.0$ all-sky quasar sample, built from \textsl{Gaia} and \textsl{unWISE} catalogs, to show that the angular distribution, redshift distribution as a function of hemisphere, amplitude of clustering as a function of sky position, and counts of pairs as a function of separation (fractal dimension), are all consistent with large-scale isotropy and homogeneity at the precision possible given the size of the catalog and the amplitude of the large-scale structure.
Roughly speaking, the results can be summarized by saying that they show the Universe to be isotropic at the sub-percent level and homogeneous in the sense of a fractal dimension of 3 to within YYY.
There is no sign of any North--South density or power asymmetry.
\abby{Is this the same hemispherical power asymmetry claimed from the CMB?}
The isotropy results can also be interpreted as a strong test of the spectrophotometric uniformity of the \textsl{Quaia} Catalog.
\end{abstract}

@@

Another reason to make new tests of isotropy and homogeneity is that data samples keep getting larger in volume and number, permitting more and more sensitive tests.
Here we make use of the brand-new \textsl{Quaia} catalog of quasars.
This catalog is built from the ESA \textsl{Gaia} DR3 data (CITE), which includes more than a million quasars, all-sky (CITE), plus photometry from the \textsl{unWISE} catalog (CITE), which is built on the NASA \textsl{WISE} infrared imaging.
This data set is not the largest cosmological data set in terms of number of tracers, but it is the largest-ever data set in terms of comoving volume, spanning some XXX $h^{-3}\,\Gpc^3$ (comoving), depending on how you want to define the sample volume.
Recent tests of homogeneity include ZZZ and ZZZ; the samples used in those studies were smaller by a factor of QQQ and WWW (approximately) in comoving volume, relative to the quasar catalog we use here.

A final but equally important reason to test isotropy and homogeneity is that these tests provide a very strong validation of the quality and uniformity of the data.
Indeed, the first work on homogeneity with the Sloan Digital Sky Survey data (\citealt{hogg05}) was performed to test the calibration and stability of the data in preparation for the measurement of the baryon acoustic feature (\citealt{baf}).
The work presented here represents a first strong test of the calibration of the \textsl{Quaia} quasar sample and associated photometry and spectroscopy; it has implications for the accuracy and uniformity of the \textsl{Gaia} data more generally.

\section{Data}\label{sec:data}\noindent
\hogg{Make sure we cite the CU5 papers and XP papers here and anything that helped make the extragalactic and quasar samples.}

@@

\hogg{Make sure to explain somewhere how the random catalog points get redshifts!!}

@@

HOGG: \figref{fig:cumulative} strongly suggests that the fractal dimension $D$ is 3.
Our best-fit value is ... WHAT?
Get quantitative here.

@@

\section{Discussion}\label{sec:discuss}\noindent
\hogg{We found what we expected to find, but at very good precision.}

\hogg{How to summarize these results?}

\hogg{Comment on the dipole and surrounding controversy.}

\hogg{What would have been different if we had gone to the $G<20.4$ sample?}

\hogg{Note that the selection function was built by assuming isotropy. In detail, that must bias our results.}

\hogg{What to say about uniformity and calibration of Quaia?}

\begin{acknowledgements}
It is a pleasure to thank
  Anna-Christina Eilers (MIT),
  Hans-Walter Rix (MPIA), and
  Abby Williams (NYU, Caltech)
for valuable discussions.
This project was developed in part at the 2022 Gaia F\^ete, hosted by the Center for Computational Astrophysics at the Flatiron Institute.
This project was also developed in part at the Gaia Hike, a workshop hosted by the University of British Columbia and the Canadian Institute for Theoretical Astrophysics in 2022 June.

\hogg{Be sure to put in the full Gaia acknowlegement.}

\hogg{Is there a WISE or unWISE acknowlegement?}
\end{acknowledgements}
